Assessing landscape N removal in coastal New England catchments using the N-Sink approach with a new R package, nsink

### 2021-06-17 manuscript/manuscript.Rmd
Edits before sending to Chet. This is the version I sent.

---
title: 'Assessing landscape N removal in coastal New England catchments using the N-Sink approach with a new R package, nsink'
# alternative title in keeping with PeerJ: nsink: an R package for estimating and visualizing N removal from landscape N sinks
author:
- |
  Hollister. J. W. ^\*^ *^1^*, Kellogg, D. Q. *^2^*, Arnold, C.*^3^*, Gold,
  A.*^2^*, Wilson, E. H.*^3^*, Chadwick, C.*^3^*, Dickson, D.*^3^*, Forshay, K.
  J.*^4^*
- |
  *^1^* US Environmental Protection Agency, Office Of Research and Development,
  Atlantic Coastal Environmental Sciences Division, Narragansett, RI 02882
- |
  *^2^* University of Rhode Island, Department of Natural Resources Science,
  Kingston, RI 02881
- |
  *^3^* University of Connecticut, Center for Land Use Education and Research,
  Haddam, CT 06438
- |
  *^4^* US Environmental Protection Agency, Office Of Research and Development,
  Groundwater Characterization and Remediation Division, Ada, OK 74820
- |
  '^\*^ \*corresponding author: <qkellogg@uri.edu*>'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: console
csl: peerj.csl
abstract: Excessive nitrogen (N) loading to coastal ecosystems impairs estuarine water quality. Land management decisions made within estuarine watersheds have a direct impact on downstream N delivery. Natural features within watersheds can act as landscape sinks for N -- wetlands, streams and ponds -- transforming dissolved N into gaseous N, effectively removing it from the aquatic system. Identifying and evaluating these landscape sinks and their spatial relationship to N sources can assist managers to evaluate the effects of their decisions on downstream resources. N-Sink is an approach that uses widely available GIS data to identify landscape sinks within HUC-12 catchments, estimate their N removal potential and map the effect of those sinks on N movement through the catchment. A series of static maps are produced to visualize N removal efficiency, N loading, transport and delivery, the latter three in the form of indices, To facilitate data acquisition, processing and visualization, an R package `nsink` was developed and has been used to produce static maps for all HUC-12 catchments in southern New England. Users can use the R package to investigate a HUC-12 of interest, or can visit the UConn website to explore the already mapped areas. Users can also investigate specific flowpaths interactively by clicking on any location within the catchment. A flowpath is generated, along with a table describing N removal along each segment.

---

```{r setup, include=FALSE}
source("../R/packages.R")
opts_chunk$set(echo = TRUE, fig.width=6, fig.height=5)
tab_num <- captioner(prefix = "Table")
fig_num <- captioner(prefix = "Figure")
options(scipen = 2, digits = 5)
```

@@

Excess nitrogen (N) delivery via surface water to coastal estuaries contributes to impaired water quality evidenced by excess algal blooms and hypoxia [@ryther1971nitrogen; @howarth2006nitrogen; @nixon1995coastal]. Hydrologic connections and flowpaths are influential in the delivery of nutrients to estuaries [@mengistu2020wetland]. Identifying landscape N sinks along hydrologic flowpaths -- areas where dissolved N is transformed to gaseous N and effectively removed from the aquatic system -- and their effect on N delivery at the watershed scale is helpful to watershed managers, land use planners and conservation organizations. The intent of N-Sink, and its associated R package `nsink`, is as a screening tool to identify areas where downgradient N sinks are able to process excess N vs. areas where downgradient N sinks are less prevalent or effective at removing N. The latter argues for more aggressive N management at the source, while the former argues for protecting and/or restoring downgradient N sinks that are helping to reduce N delivery to estuaries. N-Sink was developed as a web-based tool to visualize and explore landscape N sinks at the HUC-12 scale, and makes extensive use of widely available GIS data. The theoretical underpinnings are outlined in [@kellogg2010geospatial] and build on peer-reviewed meta-analyses and reviews [i.e., @mayer2007meta; @alexander2007role; @seitzinger2006denitrificaiton] to estimate N removal within landscape N sinks -- wetlands, streams and ponds. Residence time in these landscape features is the primary driver of N retention and transformation [e.g., @klocker2009nitrogen].

@@

Methods are what belong here and an inline fig citation looks like (`r fig_num("ww_map", display = "cite", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`) and to reference that fig, assuming it is an image file inside the ../figures folder would look this way [@bivand2014maptools].

![`r fig_num("ww_map", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`](../figures/ww_map.jpg){width="5.7in" height="7.45in"}

If you want some inline R code, do this `r nrow(iris)`.
~~~~~~

# Development of the R package `nsink`
The package `nsink` implements an approach to estimate relative nitrogen (N) removal along a flow path. This approach is detailed in Kellogg et al. [-@kellogg2010geospatial]. The `nsink` package follows from an initial version of N-Sink written in ArcGIS using ModelBuilder. That version required time-consuming data manipulation by hand due to limitations of earlier NHD datasets as well as the limitations of working in a GIS environment requiring a user license. With the increasingly more versatile GIS packages now available in R, the previously time-consuming spatial data acquisition and management of N-Sink could be automated and more easily applied to other HUC-12 catchments, leading to the development of the `nsink` package. Specifically, `nsink` relies on packages `sf` and `raster`.

# Using the `nsink` package
The `nsink` package provides several functions to set up and run an N-Sink 
analysis for a specified 12-digit HUC code.  All required data are downloaded, 
prepared for the analysis, HUC-wide nitrogen removal calculated, and flow paths 
summarized.  Additionally, a convenience function that will run all of the 
required functions for a specified HUC is included.

@@

# Download the data
`nsink_get_huc_id()` - search all 12-digit HUC names using a known location name.

@@

# Prepare the data
Once the data are downloaded there are several additional data processing steps 
that are required to subset the data to the HUC and set all data to a 
common coordinate reference system (CRS).

@@

# Calculate N removal
The next step in the N-Sink process is to calculate relative nitrogen removal.  
Details on how the nitrogen removal estimates are calculated are available in 
Kellogg et al. [-@kellogg2010geospatial]. Since the publication of that article,
changes have been implemented to make better use of the most recently available 
geospatial data; data that were not available at the time of publication. Those 
changes are described in Section XX. The `nsink_calc_removal()` function 
takes the prepared data as an input and returns a "list" with three items:

@@

# Generate and summarize flowpaths
A useful part of the N-Sink approach is not just the calculation of relative 
removal for individual components of the landscape, it is the ability to 
summarize that removal along the legnth of a specified flowpath.  The `nsink` 
package provides two functions that facilitate this process. The 
`nsink_generate_flowpath()` function takes a point location as an `sf` object 
and the prepped data (generated by `nsink_prep_data()`) as input and returns an 
`sf` LINESTRING of the flowpath starting from the input point location and 
terminating at the furthest downstream location in the input NHD Plus.  The 
flowpath on land is generated from a flow direction grid.  Once that flowpath 
intersects the surface water network, flow is determined by flow along the NHD Plus 
stream network.  First, create the `sf` POINT object.

@@

# Static maps

@@

# Convenience function: Build it all!

@@

Currently we have created the static maps for all HUC-12s in coastal southern New England (map). These may be viewed and explored interactively at https://XXXX.

# Changes to underlying methodology:

@@

# Off-network hydrology
In the NHD Plus dataset there are situations where a water body or stream segment is "off-network", i.e., it is not linked to the larger surface flow network. These may be features such as groundwater fed kettle ponds, or canals or ditches where connecting hydrology may not be captured in the NHD Plus dataset. These off-network features tend to have less NHD spatial data associated with them, making direct N removal estimates impossible. In these cases, we must estimate removal based on other in-network surface water features. While off-network features are rare and spatially small compared to the overall system, they do occur in some catchments. We estimated removal from off-network stream segments as the median removal of all first order streams in the same catchment. In this case we are assuming these first order segments behave similarly to other first order stream reaches. Off-network ditches and canals, identified as such in the NHD dataset, are expected to provide little N removal due to their designed intent to move water efficiently, reducing residence time [@buchanan2013hydrological], and therefore assign removal as the lower quartile of removal from the highest order streams in the same catchment. Finally, off-network lakes and ponds are assigned N removal at the 3rd quartile of removal from all lakes and ponds in the same catchment [@hampton2019residence; ].

# Caveats
N-Sink was designed for catchments with a surface water network, so is not intended for use in urbanized areas where surface water flow is highly manipulated and the unit of interest is a sewershed vs. a watershed. N removal from landscape sinks in urban areas would be impossible to predict. Similarly, if agricultural areas have a high density of tile drains, N removal estimates would not reflect the lower residence time and associated removal.

@@

Use Examples
===========================================================================================
N-Sink is now widely available for HUC-12 catchments in coastal CT, RI and southern MA and decision-makers are only now beginning to make use of maps and data available online (ADD URL). However, we are aware of various ways this tool is starting to be integrated into risk analyses. Here we demonstrate two examples using the Niantic River (CT) the Farm River (CT) HUCs.

@@

Results go here and if you want a table caption, do this (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1993 to 2016.")`). And then this in an example code chunk kicing out a pander table.

```{r ww_params, echo=FALSE, results="asis"}
pander::pander(head(iris), style = "rmarkdown")
```

`r tab_num("ww_params")`

>>>>

---
title: 'Assessing landscape N removal in coastal New England catchments using the N-Sink approach with a new R package, nsink'
# alternative title in keeping with PeerJ: nsink: an R package for estimating and visualizing N removal from landscape N sinks
author:
- Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Arnold, C.^*3*^, Gold,
  A.^*2*^, Wilson, E. H.^*3*^, Chadwick, C.^*3*^, Dickson, D.*^*3*^, Forshay, K.
  J.^*4*^
- ^*1*^ US Environmental Protection Agency, Office Of Research and Development,
  Atlantic Coastal Environmental Sciences Division, Narragansett, RI 02882
- ^*2*^ University of Rhode Island, Department of Natural Resources Science,
  Kingston, RI 02881
- ^*3*^ University of Connecticut, Center for Land Use Education and Research,
  Haddam, CT 06438
- ^*4*^ US Environmental Protection Agency, Office Of Research and Development,
  Groundwater Characterization and Remediation Division, Ada, OK 74820
- '^\*^ \*corresponding author: <qkellogg@uri.edu*>'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: 'Excessive nitrogen (N) loading to coastal ecosystems impairs estuarine water quality. Land management decisions made within estuarine watersheds have a direct impact on downstream N delivery. Natural features within watersheds can act as landscape sinks for N -- wetlands, streams and ponds -- transforming dissolved N into gaseous N, effectively removing it from the aquatic system. Identifying and evaluating these landscape sinks and their spatial relationship to N sources can assist managers to evaluate the effects of their decisions on downstream resources. N-Sink is an approach that uses widely available GIS data to identify landscape sinks within HUC-12 catchments, estimate their N removal potential and map the effect of those sinks on N movement through the catchment. A series of static maps are produced to visualize N removal efficiency, N loading, transport and delivery, the latter three in the form of indices, To facilitate data acquisition, processing and visualization, an R package `nsink` was developed and has been used to produce static maps for all HUC-12 catchments in southern New England. Users can use the R package to investigate a HUC-12 of interest, or can visit the UConn website to explore the already mapped areas. Users can also investigate specific flowpaths interactively by clicking on any location within the catchment. A flowpath is generated, along with a table describing N removal along each segment.'
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  cache = FALSE,
  tidy = FALSE
)
load(system.file("testdata.rda", package = "nsink"))
```

```{r setup, echo=FALSE}
library(nsink)
```

@@

Excess nitrogen (N) delivery via surface water to coastal estuaries contributes to impaired water quality evidenced by excess algal blooms and hypoxia [@ryther1971nitrogen; @howarth2006nitrogen; @nixon1995coastal]. Hydrologic connections and flowpaths are influential in the delivery of nutrients to estuaries [@mengistu2020wetland]. Identifying landscape N sinks along hydrologic flowpaths -- areas where dissolved N is transformed to gaseous N and effectively removed from the aquatic system -- and their effect on N delivery at the watershed scale is helpful to watershed managers, land use planners and conservation organizations. The intent of N-Sink, and its associated R package `nsink`, is as a screening tool to identify areas where downgradient N sinks are able to process excess N vs. areas where downgradient N sinks are less prevalent or effective at removing N. The latter argues for more aggressive N management at the source, while the former argues for protecting and/or restoring downgradient N sinks that are helping to reduce N delivery to estuaries. N-Sink was developed as a web-based tool to visualize and explore landscape N sinks at the HUC-12 scale, and makes extensive use of widely available GIS data. The theoretical underpinnings are outlined in [@kellogg2010geospatial] and build on peer-reviewed meta-analyses and reviews [i.e., @mayer2007meta; @alexander2007role; @seitzinger2006denitrification] to estimate N removal within landscape N sinks -- wetlands, streams and ponds. Residence time in these landscape features is the primary driver of N retention and transformation [e.g., @klocker2009nitrogen].

@@

## Development of the R package `nsink`
The package `nsink` implements an approach to estimate relative nitrogen (N) removal along a flow path. This approach is detailed in Kellogg et al. [-@kellogg2010geospatial]. The `nsink` package follows from an initial version of N-Sink written in ArcGIS using ModelBuilder. That version required time-consuming data manipulation by hand due to limitations of earlier NHD datasets as well as the limitations of working in a GIS environment requiring a user license. With the increasingly more versatile GIS packages now available in R, the previously time-consuming spatial data acquisition and management of N-Sink could be automated and more easily applied to other HUC-12 catchments, leading to the development of the `nsink` package. Specifically, `nsink` relies on packages `sf` and `raster`.

## Using the `nsink` package
The `nsink` package provides several functions to set up and run an N-Sink 
analysis for a specified 12-digit HUC code.  All required data are downloaded, 
prepared for the analysis, HUC-wide nitrogen removal calculated, and flow paths 
summarized.  Additionally, a convenience function that will run all of the 
required functions for a specified HUC is included.

@@

## Download the data
`nsink_get_huc_id()` - search all 12-digit HUC names using a known location name.

@@

## Prepare the data
Once the data are downloaded there are several additional data processing steps 
that are required to subset the data to the HUC and set all data to a 
common coordinate reference system (CRS).

@@

## Calculate N removal
The next step in the N-Sink process is to calculate relative nitrogen removal.  
Details on how the nitrogen removal estimates are calculated are available in 
Kellogg et al. [-@kellogg2010geospatial]. Since the publication of that article,
changes have been implemented to make better use of the most recently available 
geospatial data; data that were not available at the time of publication. Those 
changes are described in Section XX. The `nsink_calc_removal()` function 
takes the prepared data as an input and returns a "list" with three items:

@@

## Generate and summarize flowpaths
A useful part of the N-Sink approach is not just the calculation of relative 
removal for individual components of the landscape, it is the ability to 
summarize that removal along the legnth of a specified flowpath.  The `nsink` 
package provides two functions that facilitate this process. The 
`nsink_generate_flowpath()` function takes a point location as an `sf` object 
and the prepped data (generated by `nsink_prep_data()`) as input and returns an 
`sf` LINESTRING of the flowpath starting from the input point location and 
terminating at the furthest downstream location in the input NHD Plus.  The 
flowpath on land is generated from a flow direction grid.  Once that flowpath 
intersects the surface water network, flow is determined by flow along the NHD Plus 
stream network.  First, create the `sf` POINT object.

@@

## Static maps

@@

## Convenience function: Build it all!

@@

Currently we have created the static maps for all HUC-12s in coastal southern New England (map). These may be viewed and explored interactively at https://clear.uconn.edu/projects/nsink/.

## Changes to underlying methodology:

@@

## Off-network hydrology
In the NHD Plus dataset there are situations where a water body or stream segment is "off-network", i.e., it is not linked to the larger surface flow network. These may be features such as groundwater fed kettle ponds, or canals or ditches where connecting hydrology may not be captured in the NHD Plus dataset. These off-network features tend to have less NHD spatial data associated with them, making direct N removal estimates impossible. In these cases, we must estimate removal based on other in-network surface water features. While off-network features are rare and spatially small compared to the overall system, they do occur in some catchments. We estimated removal from off-network stream segments as the median removal of all first order streams in the same catchment. In this case we are assuming these first order segments behave similarly to other first order stream reaches. Off-network ditches and canals, identified as such in the NHD dataset, are expected to provide little N removal due to their designed intent to move water efficiently, reducing residence time [@buchanan2013hydrological], and therefore assign removal as the lower quartile of removal from the highest order streams in the same catchment. Finally, off-network lakes and ponds are assigned N removal at the 3rd quartile of removal from all lakes and ponds in the same catchment [@hampton2019residence; ].

## Caveats
N-Sink was designed for catchments with a surface water network. It is not intended in arid areas or where surface water flow is highly manipulated, such as in urbanized or tile-drained agricultural areas. In these situations surface water flow is highly variable and often difficult to estimate or predict.

@@

Use Examples
============
N-Sink is now available for HUC-12 catchments in coastal CT, RI and southern MA and decision-makers are beginning to make use of maps and data available as an [interactive tool](https://uconnclear.maps.arcgis.com/apps/MapSeries/index.html?appid=e96b01502f5b4aeda64aacf0cb5234c7&folderid=dde199a7cf674845a83b43a3011b1d31).
However, we are aware of various ways this tool is starting to be integrated into risk analyses. Here we demonstrate two examples using the Niantic River (CT) the Farm River (CT) HUCs.

@@
